### 2022-08-23 article.tex
Improve quote
Improve

Here we define paradata as the data that describes the process of generating data.
In genetic epidemiology, the data generated is mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, one could argue incorrectly that
it is the scientific paper describes which computation steps 
have taken place.
However, this has the unrealistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should should be supplied,
as this is the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

\epigraph{
  Talk is cheap. Show me the code. \\
  \\
  Linus Torvalds, 2000-08-25. Message to linux-kernel mailing list
}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{figure_1.png}
  \caption{
    Left: general relations between data, paradata and metadata.
    Middle and right: examples of these types of data
  }
  \label{fig:figure_1}
\end{figure}

@@

Paradata is relevant, as it allows primary data to be re-used.
In the eBird example, it helps assess the trustworthiness of the 
data, as beginner birdwatchers make more mistakes in identifying
the correct bird species, as compared to experts.
By using the name and/or experience of the birder as a covariate,
we can re-use the dataset in other contexts as well,
as now this example can be generalized to
a classification problem of unlabeled
data for learning agents with different amounts 
of supervised and unsupervised learning.

@@

This paper takes the standpoint that it is the computer code 
that describes this generation of results, 
hence it is the computer code that is paradata (i.e. not the description of the experiment),
as shown schematically in the right panel of Figure \ref{fig:figure_1}.
Reading the computer code allows one to assess the trustworthiness 
of the experiment, similar to knowing the name or expertise of a bird spotter.
Having the code allows one to re-use it, in the same or other contexts,
providing for reproducible research.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the computer code is paradata,
that this paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for 
science in general and healthcare in particular.

@@

Genetic epidemiology is a field within biology that, among 
others, measures the spread of heritable traits,
as well as how these come to be.
For example, we know that lactose intolerance is, among others,
caused by producing too few lactose-degrading enzymes,
and is most commonly found in south-east asia.
The trait is caused by the genetic make-up, or genotype, of a person.
The trait, also called phenotype, in this example is lactose intolerance,
but any human property, such as weight or height can be studied.
Genetic epidemiology studies the spread of genotypes and phenotypes,
as well as the relation between those two.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for a phenotype.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code is more important than the (English) paper}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The (scientific) article is metadata, as it is data that describes other data:
an article describes the
experiment and hence the programming
code in English. However, it is not the best
candidate to describe how the data is collected,
as it has a loose connection with collecting the data.
If the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.

@@

\begin{table}[h]
  \begin{tabular}{lp{5cm}p{5cm}}
    Term      & Definition                                            & Example                                                         \\
    \hline
    Code      & Body of text to be run by a computer                  & The scripts run by a computational experiment                   \\
    Data      & Individual facts, statistics, or items of information & A SNP that has a significant association                        \\
    Genotype  & The DNA allele at a certain location                  & AA, AC, CG, GT, ...                                             \\
    Paradata  & Data that describes the process of generating data    & The code to conclude that a SNP has a significant association   \\
    Phenotype & How an organism looks like in the broadest sense      & The concentration of IL6RA in the blood                         \\
    Primary data & Data that is worked with in a research             & The genotype of individuals   \\
    Metadata  & Data that provides information about other data       & The article that describes an experiment                        \\
    Raw data  & Data directly from the real world, that is not yet ready to be primary data & The raw results of getting the genotypes of individuals   \\
    Trait     & A phenotype                                           & The concentration of IL6RA in the blood                         
  \end{tabular}
\end{table}

>>>>

Here we define paradata as the data that describes the process of generating data.
In genetic epidemiology, the data generated is mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, one could argue incorrectly that
it is the -usually English- scientific paper that best describes which computation steps 
have taken place.
However, this has the unrealistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should should be supplied,
as this is the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

\epigraph{
  Talk is cheap. Show me the code.
}{
  Linus Torvalds, 2000-08-25
}

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{figure_1.png}
  \caption{
    Left: general relation between data, paradata and metadata.
    Middle: the same relations specified for eBird.
    Right: the same relations specified for genetic epidemiology.
  }
  \label{fig:figure_1}
\end{figure}

@@

Paradata is relevant, as it allows primary data to be re-used.
In the eBird example, it helps assess the trustworthiness of the 
data, as beginner birdwatchers tend to make more mistakes in identifying
the correct bird species, as compared to experts.
By using the name and/or experience of the birder as a covariate,
we can re-use the dataset in other contexts as well,
as now this example can be generalized to
a classification problem of unlabeled
data for learning agents with different amounts 
of experience.

@@

This paper takes the standpoint that it is the computer code 
that describes this generation of results, 
hence it is the computer code that is paradata 
(i.e. not the description of the experiment as done in a natural language),
as shown schematically in the right panel of Figure \ref{fig:figure_1}.
Reading the computer code allows one to assess the trustworthiness 
of the experiment, similar to knowing the name or expertise of a bird spotter.
Having the code allows one to 
verify, re-use and extend the experiment, 
in the same or other contexts, providing for reproducible research.
Additionnally, this paper suggests some best practices on how to make 
an experiment FAIR data (findability, accessibility, interoperability, and reusability) \cite{wilkinson2016fair} better.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the computer code is paradata,
that this paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for 
science in general and healthcare in particular.
However, any field that uses computation in its experiments
could be used as an example.

@@

Genetic epidemiology is a field within biology that, among 
others, measures the spread of heritable traits,
as well as how these come to be.
For example, we know that lactose intolerance is, among others,
caused by producing too few lactose-degrading enzymes,
and is most commonly found in south-east asia.
The trait is caused by the genetic make-up, 
or genotype (see Table \ref{tab:definitions}), of a person.
The trait, also called phenotype, in this example is lactose intolerance,
but any human property, such as weight or height can be studied.
Genetic epidemiology studies the spread of genotypes and phenotypes,
as well as the relation between those two.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for a phenotype.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code is more important than the paper}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The (scientific) article is metadata, as it is data that describes other data:
an article describes the
experiment (and hence the programming
code) in a natural language, which is usually English (at this point in time). 
However, a natural language is not the best
candidate to describe how the data is collected,
as it has a loose connection with collecting the data.
If the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.

@@

\begin{table}[h]
  \begin{tabular}{lp{5cm}p{5cm}}
    Term      & Definition                                            & Example                                                         \\
    \hline
    Code      & Body of text to be run by a computer                  & The scripts run by a computational experiment                   \\
    Data      & Individual facts, statistics, or items of information & A SNP that has a significant association                        \\
    Genotype  & The DNA allele at a certain location                  & AA, AC, CG, GT, ...                                             \\
    Paradata  & Data that describes the process of generating data    & The code to conclude that a SNP has a significant association   \\
    Phenotype & How an organism looks like in the broadest sense      & The concentration of IL6RA in the blood                         \\
    Primary data & Data that is worked with in a research             & The genotype of individuals   \\
    Metadata  & Data that provides information about other data       & The article that describes an experiment                        \\
    Raw data  & Data directly from the real world, that is not yet ready to be primary data & The raw results of getting the genotypes of individuals   \\
    Trait     & A phenotype                                           & The concentration of IL6RA in the blood                         
  \end{tabular}
  \caption{Terms used in this paper, and their definitions}
  \label{tab:definitions}
\end{table}
